---
title: "The 95% confidence interval"
author: "Sanne"
date: "23/11/2019"
output: html_document
---

```{r setup, echo=FALSE}
# Load packages
library(plotrix)
library(knitr)

# Set seed
set.seed(465458)

# Create empty vectors of length 100 for every statistic
mean_samp <- rep(0,100)
sd_samp <- rep(0,100)
```

```{r sample, echo = F}
# Draw 100 samples of 100 values
for(i in 1:100){
  samp <- rnorm(100)
  mean_samp[i] <- mean(samp) # Calculate mean of sample
  sd_samp[i] <- sd(samp) # Calculate standard deviation of sample
}
```

## Sample statistics

For this explanation, 100 x 100 samples are drawn from a normal distribution with $\mu = 0$ and $\sigma = 1$. The sample mean $\overline{x}$ and sample standard deviation $s$ are calculated for each sample. 

The **absolute bias** of the sample can be calculated using the formula $$\text{absolute bias} = abs(\overline{x} - \mu)$$.

As $\mu = 0$ in this case, $- \mu$ is omitted from the equation.

The **confidence interval** can calculated with $$CI = \overline{x} \pm z*SE$$ with $z = 1.96$ for the 95% confidence interval. $SE$ is the standard error, which is calculated using the formula 
$$SE = \frac{s}{\sqrt{n}}$$, with $n$ being the sample size.


```{r statistic, echo=FALSE}
abs_bias <- abs(mean_samp) # Absolute bias
SE <- sd_samp/sqrt(100) # Standard error
low_conf <- mean_samp - 1.96*SE # Lower bound confidence interval
high_conf <- mean_samp + 1.96*SE # Higher bound confidence interval
```

## The 95% confidence interval

The 95% confidence interval gives the range between which the true mean will lie in at least 95% of the samples. As the plot below shows, this is also true for the current sample.

```{r confint, echo=FALSE}
# Plot sample means
plot(mean_samp, ylim = c(-1,1), pch=16, ylab="Sample mean", xlab = "", xaxt = "n")
# Add a horizontal line 
abline(h=0, col="blue")
# Add confidence intervals
for(i in 1:100){
  lines(x=c(i, i), y=c(low_conf[i], high_conf[i]))
  if(low_conf[i] > 0 ||high_conf[i] < 0){
    lines(x=c(i, i), y=c(low_conf[i], high_conf[i]), col="red")
    points(i, mean_samp[i], pch=16, col="red")
  }
}
```

In this case, the confidence interval does not include the true mean in exactly 5 out of a 100 samples. For only a hundred samples, the exact number can vary due to chance. 

## Table of samples that do not include the mean
```{r table, echo=FALSE}
low <- low_conf > 0 
high <- high_conf < 0
indexes <- which(as.logical(low + high))
matrix_samp <- data.frame(mean_samp, sd_samp, low_conf, high_conf)
kable(matrix_samp[indexes,], col.names = c("Mean", "Standard deviation", "95% CI Lower", "95% CI Upper"))
```

